---
title: "Heart Failure Prediction"
date: "`r Sys.Date()`"
output: rmarkdown::github_document
---

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(hrbrthemes)
library(scales)
library(waffle)
library(ggthemes)
library(dplyr)
library(patchwork)
library(gridExtra)
library(ggpubr)
library(skimr)
library(corrplot)
library(caret) 
library(olsrr)
```

# Data importation and curation

```{r class.source = 'fold-show'}
data = read.csv("/Users/antonin/Documents/Github/Heart_Failure_Prediction_with_R/data/heart.csv", stringsAsFactors = T)
data$HeartDisease <- as.factor(data$HeartDisease)
data$FastingBS <- as.factor(data$FastingBS)
data$Cholesterol <- NULL 
``` 

The Cholesterol column bas been removed because many 0 within population with Heart disease. Biologically, a Cholesterol level cannot be equal to 0, probably NAs values that have been replaced with 0. 

To use this column next time, a benchmarking has to be done with the different models created with the following options :

  * Rows with Cholesterol = 0 removed.  
  * Replace 0 values with the mean.  
  * Replace 0 with the median.  
  * Replace 0 values using the k-mean algorithm.
  
Observe prediction results and choose the best option. 

```{r class.source = 'fold-show', fig.align = "center"}
skim(data)
```

```{r class.source = 'fold-show'}
summary(data$RestingBP)
```
Minimum 0 as repos resting mean the person is death, we can presume an anomali in the data. Let's check how many rows are concerned. 

```{r class.source = 'fold-show'}
data[data$RestingBP == 0,]
```
Only one, let's replace its value with the mean of the other row to fix that. 

```{r class.source = 'fold-show'}
data[450,"RestingBP"] <- mean(data$RestingBP[-450])
```

# Visual representation of the data 

Code of the differents plots. Click to expand. 

```{r, message=FALSE, warning=FALSE}
######################

# Age distribution by Sex 
data_summary <- data %>%
  group_by(Age, Sex) %>%
  summarise(Count = n(), .groups = 'drop')

age_distribution_sex <- ggplot(data_summary, aes(x = Age, y = ifelse(Sex == "M", -Count, Count), fill = Sex, width=1)) +
  geom_bar(stat="identity", color="#000000") +
  coord_flip() + 
  scale_y_continuous(labels = abs) + 
  labs(title = "", x = "Age", y = "Number of Patients") +
  scale_fill_manual(values = c("M" = "#399ce3", "F" = "#ff7f0e")) +
  theme_classic() + 
  guides(fill = guide_legend(reverse = TRUE, position = "bottom"))

######################

# Resting electrocardiogram results
data$RestingECG <- factor(data$RestingECG, levels = c("ST", "LVH", "Normal"))

resting_ecg_plot <- ggplot(data, aes(x=HeartDisease, fill=RestingECG)) + 
    geom_bar(color = "black", position="fill") + 
    theme_classic() + 
    scale_y_continuous(labels = scales::percent) +
    ylab("") + 
    scale_fill_manual(name = "", values = c("Normal" = "#446455", "LVH" = "#FDD262", "ST" = "#D3DDDC" ), label = c("ST" = "ST-T wave abnormality", "LVH" = "Left ventricular hypertrophy")) +
    scale_x_discrete(labels=c('No Heart Disease', 'Heart Disease')) +
    theme(axis.title.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "top") +
    guides(fill=guide_legend(nrow=3,byrow=TRUE))

######################

# Chest pain type 
data$ChestPainType <- factor(data$ChestPainType, levels = c("TA", "ATA", "NAP", "ASY"))

chest_pain_plot <- ggplot(data, aes(x=HeartDisease, fill=ChestPainType)) + 
    geom_bar(color = "black", position="fill") + 
    theme_classic() + 
    scale_y_continuous(labels = scales::percent) +
    ylab("") + 
    scale_fill_manual(name = "", values = c("ASY" = "#446455", "NAP" = "#FDD262", "ATA" = "#D3DDDC", "TA" = "#C7B19C"),
                      label = c("ASY" = "Asymptomatic", "NAP" = "Non-Anginal Pain", "ATA" = "Atypical Angina", "TA" = "Typical Angina", ""))+
    scale_x_discrete(labels=c('No Heart Disease', 'Heart Disease')) +
    theme(axis.title.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "top") +
    guides(fill=guide_legend(nrow=4,byrow=TRUE))

######################

# Resting blood pressure
custom_colors <- c("#0c7d34", "#bc5090")
restingBP_boxplot <- ggplot(data, aes(x=HeartDisease, y=RestingBP, fill = HeartDisease)) + 
    geom_boxplot() + 
    ylab("Resting blood pressure mmHg") + 
    scale_fill_manual(name = "Heart Condition", values = custom_colors, labels = c("No", "Yes")) +
    theme_classic() + 
    theme(legend.position = "bottom", 
          axis.title.x=element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank())

######################

# Maximum heart rate achieved
max_heart_rate_boxplot <- ggplot(data, aes(x=HeartDisease, y=MaxHR, fill = HeartDisease)) + 
    geom_boxplot() + 
    scale_fill_manual(name = "Heart Condition", values = custom_colors, labels = c("No", "Yes")) +
    theme_classic() + 
    ylab("Maximum heart rate achieved ") + 
    theme(legend.position = "bottom", 
          axis.title.x=element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank())

######################
# Slope of the peak exercise ST segment
st_slope_plot <- ggplot(data, aes(x=HeartDisease, fill=ST_Slope)) + 
    geom_bar(color = "black", position="fill") + 
    theme_classic() + 
    scale_y_continuous(labels = scales::percent) +
    ylab("") + 
    scale_fill_manual(name = "", values = c("Up" = "#446455", "Flat" = "#FDD262", "Down" = "#D3DDDC" ), label = c("Up" = "Upsloping", "Down" = "Downsloping")) +
    scale_x_discrete(labels=c('No Heart Disease', 'Heart Disease')) +
    theme(axis.title.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "top") +
    guides(fill=guide_legend(nrow=3,byrow=TRUE))

######################
# Fasting blood sugar
  ## No Heart Disease samples
fasting_normal <- data[data$HeartDisease == "0",]$FastingBS
fasting_normal <- data.frame("Heart Condition" = fasting_normal)

palette <- c("#0c7d34", "#bc5090")

fasting_plot_0 <- ggplot(fasting_normal, aes(x = Heart.Condition, fill = Heart.Condition)) +
  geom_bar(colour = "black") +
  theme_classic() + 
  ylim(0,375) +
  ylab("Patients with fasting blood sugar <= 120 mg/dl") +
  scale_fill_manual(name = "Heart Condition", values = palette, labels = c("No", "Yes")) +
  theme(plot.title = element_text(hjust = 0.5, size = 7),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "bottom")

  ## Heart Disease samples
fasting_disease <- data[data$HeartDisease == "1",]$FastingBS
fasting_disease <- data.frame("Heart Condition" = fasting_disease)

fasting_plot_1 <- ggplot(fasting_disease, aes(x = Heart.Condition, fill = Heart.Condition)) +
  geom_bar(colour = "black") +
  theme_classic() + 
  ylim(0,375) +
  ylab("Patients with fasting blood sugar > 120 mg/dl") +
  scale_fill_manual(name = "Heart Condition", values = palette, labels = c("No", "Yes")) +
  theme(plot.title = element_text(hjust = 0.5, size = 7),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "bottom")

# Exercice Angina
  ## No Heart Disease samples
exercice_angina_0 <- data[data$HeartDisease == "0",]$ExerciseAngina
exercice_angina_0 <- data.frame("Heart Condition" = exercice_angina_0)

palette <- c("#0c7d34", "#bc5090")

exercice_angina_plot_0 <- ggplot(exercice_angina_0, aes(x = Heart.Condition, fill = Heart.Condition)) +
  geom_bar(colour = "black") +
  ylim(0,375) +
  ylab("Patients without exercise-induced angina") + 
  theme_classic() + 
  scale_fill_manual(name = "Heart Condition", values = palette, labels = c("No", "Yes")) +
  theme(plot.title = element_text(hjust = 0.5, size = 7),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "bottom")

  ## Heart Disease samples
exercice_angina_1 <- data[data$HeartDisease == "1",]$FastingBS
exercice_angina_1 <- data.frame("Heart Condition" = exercice_angina_1)

exercice_angina_plot_1 <- ggplot(exercice_angina_1, aes(x = Heart.Condition, fill = Heart.Condition)) +
  geom_bar(colour = "black") + 
  ylim(0,375) +
  ylab("Patients with exercise-induced angina") + 
  theme_classic() + 
  scale_fill_manual(name = "Heart Condition", values = palette, labels = c("No", "Yes")) +
  theme(plot.title = element_text(hjust = 0.5, size = 7),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position = "bottom")

######################
# OldPeak - Numeric value measured in depression
palette <- c("#0c7d34", "#bc5090")

Oldpeak_0 <- data.frame(data$Oldpeak, data$HeartDisease)

Oldpeak_plot <- ggplot(data=Oldpeak_0, aes(x=data.Oldpeak, group=data.HeartDisease, fill=data.HeartDisease)) +
    geom_density(adjust=1, alpha=0.5) +
    theme_classic() + 
    scale_fill_manual(name = "Heart Condition", values = palette) + 
    labs(title = "", x = "ST depression value mesured", y = "") +
    theme(plot.title = element_text(hjust = 0.5)) + 
    scale_fill_manual(name = "Heart Condition", values = custom_colors, labels = c("No", "Yes"))

######################
# Merge plots

boxplots <- ggarrange(restingBP_boxplot, max_heart_rate_boxplot, ncol=2, nrow=1, common.legend = TRUE, legend="bottom", labels = c("B", "C")) 
fasting <- ggarrange(fasting_plot_0, fasting_plot_1, ncol=3, nrow=1, common.legend = TRUE, legend="none")
angina <- ggarrange(exercice_angina_plot_0, exercice_angina_plot_1, ncol=2, nrow=1, common.legend = TRUE, legend="right")

```

**Figures**:

- **A.**    Age distribution by sex
- **B.**    Resting blood pressure [mm Hg]
- **C.**    Maximum heart rate achieved [Beats per minute]
- **D.**    Count of patients with fasting blood sugar above and below 120 mg/dl
- **E.**    Count of patients with and without exercise-induced angina
- **F.**    Distribution of chest pain type (%)
- **G.**    Distribution of resting electrocardiogram results [Normal: Normal, ST: having ST-T wave abnormality (T wave inversions and/or ST elevation or depression of > 0.05 mV), LVH: showing probable or definite left ventricular hypertrophy by Estes' criteria] (%)
- **H.**    Distribution of the slope of the peak exercise ST segment (%)
- **I.**    Density of oldpeak ST [Numeric value measured in depression]


```{r, dpi=300}
# Adding labels
ggarrange(age_distribution_sex, boxplots, ncol = 2, nrow = 1, labels = c("A"))
ggarrange(fasting, angina, ncol=2, nrow=1, labels = c("D", "E")) 
ggarrange(chest_pain_plot, resting_ecg_plot, st_slope_plot, ncol=3, nrow=1, labels = c("F", "G", "H"))
ggarrange(Oldpeak_plot, ncol=1, nrow=1, labels = c("I"))
```


At first glance, we can clearly see a surepresentation of Male within the data, and a sample age tending toward 55 years old (Fig. A). The boxplot  showing the resting blood pressure (Fig. B) doesn't visually give the impression having a link with heart diseases, contrary to the plot of the maximum heart rate achieved (Fig. C). To confirm or infirm our observation, we can statistically prove or not the correlation. 

Before that, we have to test if the Resting blood pressure results and maximum heart rate achieved follow a normal distribution. 

```{r}
shapiro.test(data$RestingBP) 
shapiro.test(data$MaxHR)
```
Both p-value are below 0.05, so we can conclude they do not follow a normal distribution, and we will have to perform non-parametric correlation test (Spearman).

```{r}
cor.test(data$RestingBP, as.numeric(data$HeartDisease), method = "spearman", exact = FALSE)
cor.test(data$MaxHR, as.numeric(data$HeartDisease), method = "spearman", exact = FALSE)

```
The hypothesis H0 for cor.test is : There is no association between the two variables. However, in both case, p-value < 0.05, so we rejet the nul hypothesis. It can be concluded that both RestingBP and MaxHR do have a correclation with heart diseases. The correlation coefficient is around 0.12 for RestingBP, and -0.40 for MaxHR. Therefore, MaxHR has a higher impact on heart diseases than RestingBP. 

A higher amount of patients can be seen with high fasting blood sugar [>120 mg/dl] and exercice-induces angina among heart disease victims (Fig. D and Fig. E), both could be an important factor. 

From Fig. F, patients affected by heart diseases don't have any chest pain (asymptomatic) and generally a flat slope of the peak exercise (Fig. H).

Finally, Fig. I indicates that most of the patients without heart condition hasn't ST depression on their electrocardiogram results, when the ST segment is abnormally low below the baseline, it could be linked with heart diseases. 

To have a better understanding if some variable are correlated between each others, we can generate a correlation matrix and observe the correlation values between each conditions.

```{r, dpi=300}
######################
# Correlation matrix

table <- data.frame(as.numeric(data$Sex), data$Age, data$RestingBP, data$MaxHR, data$Oldpeak, as.numeric(data$FastingBS), as.numeric(data$ExerciseAngina), as.numeric(data$RestingECG))
names(table) <- c("Sex", "Age", "Resting BP", "Max HR", "Oldpeak", "FastingBS", "Exercice Angina", "Resting ECG")
red <- cor(table)
correlation_matrix <- corrplot(red, method="color", type="lower", addCoef.col = T,
         col = colorRampPalette(c("#446455", "#FDD262", "#D3DDDC"))(100),
         tl.col = "black", tl.srt = 45)

```

# Machine learning models  

## Data train & data test

```{r}
data <- data %>%
  mutate(HeartDisease = ifelse(HeartDisease == "1", TRUE, FALSE))
```

Before training any model, the dataset has to be split in two, 80% of the dataset will be used to train the model, and the other 20% will serve to test it. 

```{r}
eighty_percents <- round(nrow(data) * 0.2)  # Calculate how many sample represent 20% of the dataset
choice <-  sample(1:nrow(data), size = eighty_percents, replace = F) # Randomly selecting 20% of the number of patient

data.test <- data[choice, ] # Keep 20% of the dataset
data.train <- data[-choice,] # Keep the other 80% of the dataset
``` 

## Linear regression

```{r}
linear_reg_all <- lm(HeartDisease ~ ., data = data.train)
```

```{r}
data.test$pred = predict(linear_reg_all, data.test) > 0.5
confusion_matrix_linear_reg_all <- table(data.test$HeartDisease, data.test$pred)
(confusion_matrix_linear_reg_all[1,1] + confusion_matrix_linear_reg_all[2,2]) / sum(confusion_matrix_linear_reg_all) * 100

stats_linear_reg_all <- confusionMatrix(confusion_matrix_linear_reg_all)

stats_linear_reg_all$overall
stats_linear_reg_all$byClass
```

The best model isn't necessary the one with the best accuracy, because it's gonna be always one with all the factors, we can try to simplify as much as we can the model by removing the factor not bringing enough information to help for the prediction.

From the package `olsrr` (similar to `leaps`), the funcion ols_step_best_subset will perform all the combination and return the best predictors. The Mallows's C(p) is a good indicator to know which model choose at the end, balancing between complexity and accuracy. 

```{r}
ols_step_best_subset(linear_reg_all)
```

The best model following Mallows's C(p) include only the OldPeak factor instead of the 10 of the previous model.

```{r}
linear_reg_1 <- lm(HeartDisease ~ ST_Slope, data = data.train)

data.test$pred = predict(linear_reg_1, data.test) > 0.5
confusion_matrix_linear_reg_1 <- table(data.test$HeartDisease, data.test$pred)

stats_linear_reg_1 <- confusionMatrix(confusion_matrix_linear_reg_1)

stats_linear_reg_1$overall
stats_linear_reg_1$byClass
```
The performances statistics of the model are quite similar to the previous one while removing 9 factors. 
































